---
output: github_document
---

<!-- markdownlint-disable-file -->

<!-- README.md needs to be generated from README.Rmd. Please edit that file -->

# RobinCar2

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

<!-- badges: start -->

[![Project Status: Active – The project has reached a stable, usable state and is being actively developed.](https://www.repostatus.org/badges/latest/active.svg)](https://www.repostatus.org/#active) [![CRAN status](https://www.r-pkg.org/badges/version-last-release/RobinCar2)](https://www.r-pkg.org/badges/version-last-release/RobinCar2) [![CRAN monthly downloads](https://cranlogs.r-pkg.org/badges/RobinCar2)](https://cranlogs.r-pkg.org/badges/RobinCar2) [![CRAN total downloads](https://cranlogs.r-pkg.org/badges/grand-total/RobinCar2)](https://cranlogs.r-pkg.org/badges/grand-total/RobinCar2) [![Code Coverage](https://raw.githubusercontent.com/openpharma/RobinCar2/_xml_coverage_reports/data/main/badge.svg)](https://raw.githubusercontent.com/openpharma/RobinCar2/_xml_coverage_reports/data/main/coverage.xml)

<!-- badges: end -->

\

RobinCar2 is a package that provides robust covariate adjustment methods for estimating and inferring treatment effects under stratified randomization. The implemented methods follow the [FDA’s final guidance on covariate adjustment](https://www.regulations.gov/docket/FDA-2019-D-0934) and are supported by a well-established body of literature. For linear adjustment, see, for example, [Tsiatis (2008)](https://pubmed.ncbi.nlm.nih.gov/17960577/), [Bugni et al. (2018)](https://pubmed.ncbi.nlm.nih.gov/30906087/), [Ye, Shao, Yi, and Zhao (2023)](https://doi.org/10.1080/01621459.2022.2049278), and [Ye, Shao, and Yi (2022)](https://doi.org/10.1093/biomet/asab015). For nonlinear adjustment, see, e.g., [Rosenblum and van der Laan (2010)](https://www.degruyter.com/document/doi/10.2202/1557-4679.1138/html?lang=en&srsltid=AfmBOoo_ikKbaLeBYcxie-99GCMnZfRNzf6aGn8Uxxnw0X3boYtm69U0), [Wang et al. (2021)](https://doi.org/10.1080/01621459.2021.1981338), [Ye, Bannick, Yi, and Shao (2023)](https://doi.org/10.1080/24754269.2023.2205802), and [Bannick, Shao, Liu, Du, Yi, and Ye (2024)](https://arxiv.org/abs/2306.10213).

All the following papers (added above) are cited in the FDA guidance.

-   Bugni, F, IA Canay, and AM Shaikh, 2018, Inference Under Covariate-Adaptive Randomization, Journal of the American Statistical Association, 113(524):1784-1796.

-   Tsiatis, AA, M Davidian, M Zhang, and X Lu, 2008, Covariate Adjustment for Two-Sample Treatment Comparisons in Randomized Trials: A Principled Yet Flexible Approach, Statistics in Medicine, 27(23):4658-4677.

-   Wang, B, R Susukida, R Mojtabai, M Amin-Esmaeili, and M Rosemblum, 2021. Model-Robust Inference for Clinical Trials that Improve Precision by Stratified Randomization and Covariate Adjustment, Journal of the American Statistical Association, doi: 10.1080/01621459.2021.1981338.

-   Rosenblum, M and MJ van der Laan, 2010, Simple, Efficient Estimators of Treatment Effects in Randomized Trials Using Generalized Linear Models to Leverage Baseline Variables, International Journal of Biostatistics, 6(1):13.

## Installation

### Development

You can install the current development version from `github` with:

```{r gh-installation, eval = FALSE}
if (!require("remotes")) {
  install.packages("remotes")
}

remotes::install_github(
  "openpharma/RobinCar2"
)
```

## Citing `RobinCar2`

To cite `RobinCar2` please see [here](https://openpharma.github.io/RobinCar2/main/authors.html#citation).

## Acknowledgement

This package is supported by [ASA-BIOP Covariate Adjustment SWG Software Subteam](https://carswg.github.io/subteam_software.html)
