---
title: "Design Doc for Survival Analysis with RobinCar2"
author: Daniel Sabanes Bove
bibliography: references.bib
number-sections: true
---

This design doc shall serve to align on the design of the survival analysis implementation in RobinCar2.

# User interface

The proposal for the user interface is the function call of the form:

```{r}
#| eval: false

robin_surv(
    Surv(time, event) ~ treatment * s1 + covariates,
    data = df,
    treatment = pb(s1),
    contrast = "hazardratio",
    pval = "logrank"
)
```

## Name and arguments

- The name `robin_surv()` indicates general survival analysis functionality. In particular, we do not need to assume a Cox proportional hazards model in order to interpret the logrank test results. 
- We use the standard `Surv(time, event)` syntax from the `survival` package, with which users are familiar already.
    - If the user would not like to adjust for covariates, they can omit the `covariates` in the formula.
    - If the user would not like to have a stratified estimate, they can omit the `s1` in the formula and in the `treatment` argument.
- We keep in line with the other RobinCar2 functions for specifying the randomization scheme via the `treatment` argument.
- For the `contrast` argument, we will start only with the hazard ratio option, as laid out in the paper by @YeShaoYi2023. 
    - In the future, this could include other contrasts, such as differences of restricted mean survival times or differences of survival probabilities at a given time point.
- The `pval` argument allows the user to specify the type of p-value they want to compute. We will start with the logrank test, but in the future we could add other options, such as p-values based on the Cox score test, which is available in the RobinCar package already.
    - This means we always both estimate the hazard ratio and perform the corresponding logrank test.

## Output

This could be the output from the `robin_surv()` function:

````
#> Model        :  Surv(time, event) ~ treatment * s1 + covariates
#> Randomization:  treatment ~ pb(s1)  ( Permuted-Block )
#> 
#> Contrast     :  Hazard ratio
#> Test         :  Logrank
#>                Estimate Std.Err Z Value  Pr(>|z|)    
#> trt1 v.s. pbo   0.56365 0.10074  5.5952 2.203e-08 ***
#> trt2 v.s. pbo   0.77093 0.10133  7.6082   0.05235 . 
#> trt2 v.s. trt1  0.20728 0.10683  1.9402 2.779e-14 *** 
#> ---
#> Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
````

Please note that for now there will be no "Marginal Mean" section included in the output. 
However, this could be added in the future, see @sec-questions below.

# Internal implementation

The `robincar_surv()` will follow a similar structure as e.g. `robincar_glm()`. However, the methods are quite different, e.g. here we do not have the prediction of counterfactuals. Therefore the pattern still looks quite different.

Steps include:

1. Checks
1. For each pair of treatment arms, we will compute:
   a. hazard ratio estimate (incl. standard error)
   a. p-value for the logrank test
   The details will depend on the stratification and covariate adjustment choices of the user.
1. Return the list with results, with class `surv_effect`

Note that the class of the value is not `treatment_effect` here, because we don't have marginal means, or Jacobian matrix here, in contrast to GLMs. Since there is some overlap, which could e.g. be used for the `print()` method or others, it could be helpful to have a common class `treatment_effect` and then have subclasses `glm_effect` and `surv_effect` which contain additional information.

# Questions {#sec-questions}

1. For the future, which kind of marginal means could we display? E.g. restricted mean survival times which are also covariate and stratification adjusted would be great. 
    - This would then eventually parallel the results we obtain for GLMs
    - Does the methodology exist already? E.g. from references in @MagirrWangDengMorrisBaillie2024 ?
    - Note that this can come later of course. Was just an early idea 🙂 
1. If we have more than two treatment arms in the data set, we would currently just compute the pairwise contrasts. 
    - Is there a more efficient way to compute the pairwise hazard ratios and tests maybe?
    - Is this sufficient? Or do we need to / could we compute an overall (logrank) test as well?    
1. How should ties be handled?
    - The "Breslow approximation" seems simplest, see e.g. in the `survival` package's manual:

        "The Breslow approximation in essence ignores the ties. If two subjects A and B died on day 100, say, the partial likelihood will have separate terms for the two deaths. Subject A will be at risk for the death that happened to B, and B will be at risk for the death that happened to A. In life this is not technically possible of course: whoever died first will not be at risk for the second death."
    - So we use just Breslow for now.
1. If a stratum only contains a single patient, do we count this patient in `n`?
    - Question arises because this stratum then does not contribute to the statistics (summand is 0).
    - Seems we still need to count it, at least like this we match the `survival` package's result exactly.
1. In the covariate adjusted hazard ratio estimate score function, should the regression coefficients really be fixed given the original estimate? I expected them to depend on the parameter $\vartheta$ as well, but both in the paper as well as in the supplementary material they are fixed at $\widehat{\beta}_{\text{L}}$ and $\widehat{\gamma}_{\text{SL}}$ respectively.
    - Has been clarified, basically fixing the regression coefficients is a more approximate version, which is computationally simpler. We can provide both options e.g. via some expert `control` argument.

# Prototype

Here we will now include a more package-ready prototype of the survival analysis implementation, which avoids duplicate code and implements also the above described user interface. Please see the initial design doc [here](survival_initial_code.qmd) for all the details as a starting point when reading.

## Example data

As example data, also here we are going to use the NCCTG Lung Cancer Data set, which is available in the `survival` package.
We are going to use the male patients as treatment arm $j = 0$ and the female patients as treatment arm $j = 1$.

```{r}
library(dplyr)
library(checkmate)
dat <- survival::lung |>
    mutate(
        status = factor(ifelse(status == 1, "Alive", "Dead")),
        sex = factor(ifelse(sex == 1, "Male", "Female")),
        status_numeric = as.numeric(status == "Dead"),
        strata = factor(ph.ecog)
    )
head(dat)
```

# References
