
---
title: "Package Introduction"
package: RobinCar2
output:
  rmarkdown::html_vignette:
          toc: true
vignette: |
  %\VignetteIndexEntry{Package Introduction}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

RobinCar2 provides robust inference on treatment effect estimates obtained through
(generalized) linear models(glm) with variance link functions.

Here with the dummy data provided in the package, we demonstrate how to use the package.

# Common Usage

```{r}
library(RobinCar2)
head(dummy_data)
```

## Data Introduction

In the `dummy_data`, we have the following columns:

1. `id` is the patient identifier.
1. `treatment` is the treatment assignment.
1. `s1` is the first stratification factor.
1. `s2` is the second stratification factor.
1. `covar` is the continuous covariate.
1. `y` is the continuous outcome.
1. `y_b` is the binary outcome.

## Obtain Treatment Effect for Continuous Outcome

In simplest case of a continuous outcome, and we want to adjust for `covar`, `s1`, where `covar` is
a covariate and `s1` is a stratification factor.
Randomization is permuted-block randomization stratified by `s1`.
A proper model to use is `y ~ treatment * s1 + covar`.

```{r}
robin_glm(y ~ treatment * s1 + covar, data = dummy_data, treatment = treatment ~ pb(s1))
```

We can also use the Huber-White variance estimator by setting `vcov = "vcovHC"`.
Please note that in this case, the model formula should not contain treatment interaction with stratification factors.

```{r}
robin_glm(y ~ treatment + s1 + covar, data = dummy_data, treatment = treatment ~ pb(s1), vcov = "vcovHC")
```

## Obtain Treatment Effect for Binary Outcome

If the outcome is binary, and we want to adjust for `covar`, `s1`, where `covar` is
a covariate and `s1` is a stratification factor.
Randomization is permuted-block randomization stratified by `s1`.
A proper model to use is `y_b ~ treatment * s1 + covar`.
Note here we need to specify `family` to be `binomial(link = "logit")`.

```{r}
robin_glm(y_b ~ treatment * s1 + covar, data = dummy_data, treatment = treatment ~ pb(s1), family = binomial(link = "logit"))
```

## Obtain Treatment Effect for Counts

If the outcome is count, and we want to adjust for the same covariates,
we can use `family = MASS::negative.binomial(theta = NA)`, to correctly optimize over `theta`.
Otherwise we can also have a fixed `theta` if we do know the truth.

Other than negative binomial link function, it is also possible that we use Poisson link function.

```{r}
dummy_data$y_count <- rpois(nrow(dummy_data), lambda = 20)
robin_glm(
  y_count ~ treatment * s1 + covar,
  data = dummy_data,
  treatment = treatment ~ pb(s1), family = MASS::negative.binomial(theta = 1)
)
```

## Using Different Covariate-Adaptive Randomization Schema

If the randomization schema is not permuted-block randomization, we can use other randomization schema.
Currently we have also `sp`(simple), `pb`(permuted-block), `ps`(Pocock-Simon).

```{r}
robin_glm(y_b ~ treatment * s1 + covar, data = dummy_data, treatment = treatment ~ ps(s1), family = binomial(link = "logit"))
```
