
---
title: "Package Introduction"
package: RobinCar2
output:
  rmarkdown::html_vignette:
          toc: true
vignette: |
  %\VignetteIndexEntry{Package Introduction}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

RobinCar2 provides robust covariate adjustment methods for estimating and
inferring treatment effects through generalized linear models (glm) under
different randomization schema.

# Common Usage
A minimal call of `robin_lm()` and `robin_glm()`, consisting of only formula,
data arguments and the randomization scheme, will produce an object of class
`treatment_effect`.

```{r}
library(RobinCar2)
head(glm_data)
```

## Data Introduction

In the `glm_data`, we have the following columns:

1. `id` is the patient identifier.
1. `treatment` is the treatment assignment.
1. `s1` is the first stratification factor.
1. `s2` is the second stratification factor.
1. `covar` is the continuous covariate.
1. `y` is the continuous outcome.
1. `y_b` is the binary outcome.

## Obtain Treatment Effect for Continuous Outcomes Using the General Variance

For the continuous outcome `y`, the linear model includes `covar` as a covariate,
`s1` as a stratification factor. The randomization scheme is a permuted-block
randomization stratified by `s1`. The model formula also includes the treatment
by stratification interaction as `y ~ treatment * s1 + covar`.

```{r}
robin_lm(y ~ treatment * s1 + covar,
  data = glm_data,
  treatment = treatment ~ pb(s1)
)
```

We can also use the Huber-White variance estimator by setting `vcov = "vcovHC"`.
Please note that in this case, the model formula should not contain the
treatment by stratification (covariate) interaction.

```{r}
robin_lm(y ~ treatment + s1 + covar,
  data = glm_data,
  treatment = treatment ~ pb(s1),
  vcov = "vcovHC"
)
```

Note that `robin_glm` can also handle continuous outcomes using the default
family and the default link function `family = gaussian()`.

## Obtain Treatment Effect for Binary Outcomes

For binary outcomes, the logistic model includes `covar` as a covariate,
`s1` as a stratification factor. The randomization scheme is a permuted-block
randomization stratified by `s1`. The model formula also includes the treatment
by stratification interaction as `y_b ~ treatment * s1 + covar`. Note here we need
to specify `family` to be `binomial(link = "logit")`.

```{r}
robin_glm(y_b ~ treatment * s1 + covar,
  data = glm_data,
  treatment = treatment ~ pb(s1),
  family = binomial(link = "logit")
)
```

## Obtain Treatment Effect for Counts

For counts, the log link model includes `covar` as a covariate,
`s1` as a stratification factor. The randomization scheme is a permuted-block
randomization stratified by `s1`. The model formula also includes the treatment
by stratification interaction as `y_count ~ treatment * s1 + covar`. Note here
we need to specify `family` to be `poisson(link = "log")` to use the Poisson
model or to be `MASS::negative.binomial(theta = NA)` to use the negative
binomial model. A fixed `theta` could be provided if it is known.

```{r}
glm_data$y_count <- rpois(nrow(glm_data), lambda = 20)
robin_glm(
  y_count ~ treatment * s1 + covar,
  data = glm_data,
  treatment = treatment ~ pb(s1),
  family = MASS::negative.binomial(theta = 1)
)
```

## Using Different Covariate-Adaptive Randomization Schema

If the randomization schema is not permuted-block randomization, we can use
other randomization schema. Currently RobinCar2 supports `sr` for the simple
randomization (this was previously called `sp`), `pb` for the permuted-block 
randomization, and `ps` for the Pocock-Simon randomization.

```{r}
robin_glm(y_b ~ treatment * s1 + covar,
  data = glm_data,
  treatment = treatment ~ ps(s1),
  family = binomial(link = "logit")
)
```

## How the Randomization Scheme Enters Each Function

All analysis functions give valid inference under covariate-adaptive
randomization, but they get there by two different routes, and the route
determines whether the randomization strata have to appear in the analysis
formula.

`robin_lm()` and `robin_glm()` read the randomization strata from the
`treatment` formula, *independently of the model formula*, and use them to build
a scheme-specific variance correction (zero under `sr` and `ps`, non-zero under
`pb`). So a randomization variable may be omitted from the model without
invalidating anything -- here we randomize on `s1` and `s2` but adjust for
neither:

```{r}
robin_glm(y_b ~ treatment,
  data = glm_data,
  treatment = treatment ~ pb(s1, s2),
  family = binomial(link = "logit")
)
```

No warning: the permuted-block dependence is still accounted for. Omitting `s1`
and `s2` costs precision, not validity.

`robin_mh()` and `robin_surv()` have no such correction term. They are
scheme-agnostic precisely *because* they condition on the strata directly, which
leaves no residual scheme-induced dependence to correct. The analysis formula is
therefore the only route by which randomization is accounted for, so dropping a
randomization variable is flagged:

```{r}
robin_mh(y_b ~ s1,
  data = subset(glm_data, treatment != "trt2"),
  treatment = treatment ~ pb(s1, s2)
)
```

The direction of the failure is worth being precise about. Permuted blocks force
treatment balance within each stratum, which *removes* variability that the
unstratified formula still assumes is present, so the standard error comes out
too large: the interval over-covers and power is lost, rather than false
positives being produced. In a 2000-replicate simulation (n = 400, permuted
blocks within a stratum with risks 0.15 and 0.65, true risk difference 0.10),
`y_b ~ 1` gave a mean standard error 18% above the true sampling SD and 98%
coverage of a nominal 95% interval, while `y_b ~ s1` gave a ratio of 1.02 and
95.0% coverage. Under simple randomization there is no such penalty -- `sr`
forces no balance, so the unstratified standard error is already correct
(ratio 0.99, coverage 95.2%); stratifying then buys precision, not correctness.

| | Randomization handled by | Randomization variable missing from the analysis formula |
|---|---|---|
| `robin_lm()`, `robin_glm()` | scheme-aware variance correction, fed from `treatment` | costs precision only; no warning |
| `robin_mh()`, `robin_surv()` | conditioning on the strata in the estimator | conservative standard error under `pb`; warning issued |

One case is a genuine bias rather than a precision loss: if the allocation ratio
varies across the omitted variable, the arms differ in composition and the
unstratified estimate is confounded. With 150/50 patients in one stratum and
50/150 in the other and a true risk difference of 0.10, `y_b ~ 1` returns 0.375
while `y_b ~ s1` returns 0.15. Stratified randomization rules this out by
construction, which is one more reason to keep the randomization strata in the
formula.

This is also why the scheme wrapper (`pb`, `ps`, `sr`) changes the numbers for
`robin_lm()` and `robin_glm()` but is only informational for `robin_mh()` and
`robin_surv()`. For the trade-offs specific to each -- how fine to make the
strata, and which variables belong there -- see `?robin_mh` and the
"Survival Analysis with RobinCar2" article.

## Obtain the Confidence Intervals for the Marginal Means and contrast

To obtain the confidence interval, we can use `confint` function.

Given the following model
```{r}
robin_res <- robin_glm(y_b ~ treatment * s1 + covar,
  data = glm_data,
  treatment = treatment ~ ps(s1),
  family = binomial(link = "logit"),
  contrast = "log_risk_ratio"
)
robin_res
```

It is easy to obtain the confidence interval matrix for the marginal mean, at specified level.
If `parm` is not provided, the complete matrix (all treatment groups) will be provided.

```{r}
confint(robin_res$marginal_mean, parm = 1:2, level = 0.7)
confint(robin_res$marginal_mean, level = 0.7)
```

Similarly for the contrast, however it has an additional argument `transform` to provide the confidence interval at transformed level.
Thus, standard error is removed because it does not make sense anymore.
By default, if the `log_risk_ratio` or `log_odds_ratio` is used as contrast, the `confint` will transform it back using exponential function.
You can also specify the `transform` to be `identity` to avoid the transformation.

```{r}
confint(robin_res$contrast)
confint(robin_res$contrast, transform = identity)
```
