---
title: "RobinCar2"
date: "`r Sys.Date()`"
author: Liming Li
---

#  Needed Methodology

Per discussion, we definitely need the glm (including lm) as our priority.
So here we will focus on the implementation of glm methods.

# Prototypes

## Using a new fit function

In this proposal, we can use the following function as our interface

```{r, eval = FALSE}
fit <- robin_glm(y ~ trt + x1 + x2, ..., treat = "trt", car_scheme = simple())
treatment_effect(fit, contrast = difference)
```

## Using a fitted object

For a fitted object, it is easy to obtain its data and model.
The method can be easier to use.

```{r, eval = FALSE}
fit <- glm(y ~ trt + x1 + x2, ...)
summary_measure <- robin(fit, treat = "trt", car_scheme = simple())
treatment_effect(summary_measure)
```


Note: we can actually enable all? e.g.

```{r, eval = FALSE}
robin_glm <- function(...) {
  fit <- glm(...)
  robin(fit, ...)
}
```

## Prototypes and benchmarking

```{r}
library(RobinCar)
library(covadj)
library(waldo)
library(microbenchmark)
```

### Anova

```{r}
fit <- lm(y ~ A, data = data)

robin <- robincar_linear(df = data, 
                             response_col="y",
                             treat_col="A",
                             covariate_cols=c("x1", "x3"),
                             car_scheme="simple",
                             adj_method="ANOVA")
compare(vcov_robin(fit, "A"), robin$varcov, ignore_attr = TRUE, tolerance = 1e-7)


microbenchmark(
  a1 = {fit <- lm(y ~ A, data = data); vcov_robin(fit, "A")},
  a2 = {fit <- glm(y ~ A, data = data); vcov_robin(fit, "A")},
  b = {robin <- robincar_linear(df = data, 
                             response_col="y",
                             treat_col="A",
                             covariate_cols=c("x1", "x3"),
                             car_scheme="simple",
                             adj_method="ANOVA"); robin$varcov},
  check = "equivalent"
)
```

### Ancova

```{r}
fit <- lm(y ~ A + x1 + x3, data = data)

robin <- robincar_linear(df = data, 
                            response_col="y",
                            treat_col="A",
                            strata_cols=c("z1", "z2"),
                            covariate_cols=c("x1", "x3"),
                            car_scheme="biased-coin",
                            adj_method="ANCOVA")

compare(vcov_robin(fit, "A", strata_formula = ~ A + z1 + z2), robin$varcov, ignore_attr = TRUE, tolerance = 1e-7)

microbenchmark(
  a1 = {fit <- lm(y ~ A + x1 + x3, data = data); vcov_robin(fit, "A", strata_formula = ~ A + z1 + z2)},
  a2 = {fit <- glm(y ~ A + x1 + x3, data = data); vcov_robin(fit, "A", strata_formula = ~ A + z1 + z2)},
  b = {robin <- robincar_linear(df = data, 
                            response_col="y",
                            treat_col="A",
                            strata_cols=c("z1", "z2"),
                            covariate_cols=c("x1", "x3"),
                            car_scheme="biased-coin",
                            adj_method="ANCOVA"); robin$varcov},
  check = "equivalent"
)
```

### ANHECOVA

```{r}
fit <- lm(y ~ A + A:x1 + A:x3 + A:z1:z2, data = data)
robin <- robincar_linear(df = data, 
                                response_col="y",
                                treat_col="A",
                                strata_cols=c("z1", "z2"),
                                covariate_cols=c("x1", "x3"),
                                car_scheme="pocock-simon",
                                adj_method="ANHECOVA",
                                contrast_h=NULL)
compare(vcov_robin(fit, "A"), robin$varcov, ignore_attr = TRUE, tolerance = 1e-7)

microbenchmark(
  a1 = {fit <- lm(y ~ A + A:x1 + A:x3 + A:z1:z2, data = data); vcov_robin(fit, "A")},
  a2 = {fit <- glm(y ~ A + A:x1 + A:x3 + A:z1:z2, data = data); vcov_robin(fit, "A")},
  b = {robin <- robincar_linear(df = data, 
                                response_col="y",
                                treat_col="A",
                                strata_cols=c("z1", "z2"),
                                covariate_cols=c("x1", "x3"),
                                car_scheme="pocock-simon",
                                adj_method="ANHECOVA",
                                contrast_h=NULL); robin$varcov},
  check = "equivalent"
)

```

### Glm 

```{r}
fit <- glm(y_bin ~ A + A:z1, data = data,  family = binomial(link = "logit"))

robin <- robincar_glm(df = data,
                              response_col="y_bin",
                              treat_col="A",
                              strata_cols=c("z1"),
                              car_scheme="pocock-simon",
                              g_family=binomial(link="logit"),
                              g_accuracy=7,
                              adj_method="heterogeneous",
                              covariate_to_include_strata=TRUE)

compare(vcov_robin(fit, "A"), robin$varcov, ignore_attr = TRUE, tolerance = 1e-7)

microbenchmark(
  a1 = {fit <- glm(y_bin ~ A + A:z1, data = data,  family = binomial(link = "logit")); vcov_robin(fit, "A")},
  b = {robin <- robincar_glm(df = data,
                              response_col="y_bin",
                              treat_col="A",
                              strata_cols=c("z1"),
                              car_scheme="pocock-simon",
                              g_family=binomial(link="logit"),
                              g_accuracy=7,
                              adj_method="heterogeneous",
                              covariate_to_include_strata=TRUE); robin$varcov},
  check = "equivalent"
)
```

### glm hetero2

```{r}
fit <- glm(y_bin ~ A + A:x1 + A:z2, data = data,  family = binomial(link = "logit"))

robin <- robincar_glm(df = data,
                              response_col="y_bin",
                              treat_col="A",
                              strata_cols=c("z2"),
                              covariate_cols=c("x1"),
                              car_scheme="biased-coin",
                              g_family=binomial(link="logit"),
                              g_accuracy=7,
                              covariate_to_include_strata=TRUE,
                              adj_method="heterogeneous")

compare(vcov_robin(fit, "A"), robin$varcov, ignore_attr = TRUE, tolerance = 1e-7)

microbenchmark(
  a1 = {fit <- glm(y_bin ~ A + A:x1 + A:z2, data = data,  family = binomial(link = "logit")); vcov_robin(fit, "A")},
  b = {robin <- robincar_glm(df = data,
                              response_col="y_bin",
                              treat_col="A",
                              strata_cols=c("z2"),
                              covariate_cols=c("x1"),
                              car_scheme="biased-coin",
                              g_family=binomial(link="logit"),
                              g_accuracy=7,
                              covariate_to_include_strata=TRUE,
                              adj_method="heterogeneous"); robin$varcov},
  check = "equivalent"
)
```

### glm hetero3

```{r}
fit <- glm(y_bin ~ A + z2:x1, data = data, family = binomial(link = "logit"))

robin <- robincar_glm(df = data,
                            response_col="y_bin",
                            treat_col="A",
                            formula="y_bin ~ A + z2:x1",
                            strata_cols=c("z2"),
                            car_scheme="biased-coin",
                            g_family=binomial(link="logit"),
                            g_accuracy=7)

compare(vcov_robin(fit, "A", strata_formula = ~A + z2), robin$varcov, ignore_attr = TRUE, tolerance = 1e-7)

microbenchmark(
  a1 = {fit <- glm(y_bin ~ A + z2:x1, data = data, family = binomial(link = "logit")); vcov_robin(fit, "A", strata_formula = ~A + z2)},
  b = {robin <- robincar_glm(df = data,
                            response_col="y_bin",
                            treat_col="A",
                            formula="y_bin ~ A + z2:x1",
                            strata_cols=c("z2"),
                            car_scheme="biased-coin",
                            g_family=binomial(link="logit"),
                            g_accuracy=7); robin$varcov},
  check = "equivalent"
)
```

### glm hetero4

```{r}
fit <- glm(y_bin ~ A + A:x1 + A:z2, data = data, family = binomial(link = "logit"))
robin <- glm.heterogeneous <- robincar_glm(df = data,
                                  response_col="y_bin",
                                  treat_col="A",
                                  strata_cols=c("z2"),
                                  covariate_cols=c("x1"),
                                  car_scheme="biased-coin",
                                  g_family=binomial(link="logit"),
                                  g_accuracy=7,
                                  covariate_to_include_strata=TRUE,
                                  adj_method="heterogeneous",
                                  contrast_h=NULL)

compare(vcov_robin(fit, "A"), robin$varcov, ignore_attr = TRUE, tolerance = 1e-7)

microbenchmark(
  a1 = {fit <- glm(y_bin ~ A + A:x1 + A:z2, data = data, family = binomial(link = "logit")); vcov_robin(fit, "A", strata_formula = ~A + z2)},
  b = {robin <- glm.heterogeneous <- robincar_glm(df = data,
                                  response_col="y_bin",
                                  treat_col="A",
                                  strata_cols=c("z2"),
                                  covariate_cols=c("x1"),
                                  car_scheme="biased-coin",
                                  g_family=binomial(link="logit"),
                                  g_accuracy=7,
                                  covariate_to_include_strata=TRUE,
                                  adj_method="heterogeneous",
                                  contrast_h=NULL); robin$varcov},
  check = "equivalent"
)
```