---
title: "Mantel-Haenszel Risk Difference in RobinCar and RobinCar2"
output: rmarkdown::html_document
vignette: >
  %\VignetteIndexEntry{Mantel-Haenszel Risk Difference in RobinCar and RobinCar2}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
if (!requireNamespace("RobinCar", quietly = TRUE)) {
  install.packages("RobinCar")
}
library(RobinCar2)
library(RobinCar)
```

`RobinCar2` provides the Mantel-Haenszel (MH) risk difference and the average
treatment effect (ATE) on the risk-difference scale via `robin_mh()`. The
underlying estimands and variance estimators match `RobinCar::robincar_mh()`,
but the package grammar matches the rest of `RobinCar2`: the response,
analysis strata, treatment column and randomization scheme are conveyed via
two formulas.

## Interface

```{r}
df_two <- subset(glm_data, treatment != "trt2")
df_two$treatment <- droplevels(df_two$treatment)
```

### `RobinCar::robincar_mh()`

```{r}
robincar_mh(
  df = df_two,
  treat_col = "treatment",
  response_col = "y_b",
  strata_cols = c("s1", "s2"),
  estimand = "ATE",
  ci_type = "mGR"
)
```

### `RobinCar2::robin_mh()`

```{r}
robin_mh(
  formula = y_b ~ s1 + s2,
  data = df_two,
  treatment = treatment ~ pb(s1, s2),
  estimand = "ATE",
  ci_type = "mGR"
)
```

The two estimators agree to machine precision; `RobinCar2` reports the contrast
as `trt1 v.s. pbo`, matching the sign convention used elsewhere in the package
(experimental arm minus reference arm).

## Numerical equivalence across all estimand / variance combinations

```{r}
specs <- expand.grid(
  estimand = c("MH", "ATE"),
  ci_type = c("GR", "mGR", "Sato"),
  stringsAsFactors = FALSE
)
specs <- specs[!(specs$estimand == "ATE" & specs$ci_type != "mGR"), ]

cmp <- do.call(rbind, lapply(seq_len(nrow(specs)), function(i) {
  est <- specs$estimand[i]
  ci <- specs$ci_type[i]
  new <- robin_mh(
    y_b ~ s1 + s2,
    data = df_two,
    treatment = treatment ~ pb(s1, s2),
    estimand = est, ci_type = ci
  )
  ref <- robincar_mh(
    df = df_two, treat_col = "treatment", response_col = "y_b",
    strata_cols = c("s1", "s2"), estimand = est, ci_type = ci
  )$result
  data.frame(
    estimand   = est,
    ci_type    = ci,
    new_est    = unname(new$estimate),
    ref_est    = -as.numeric(ref$estimate), # sign convention
    new_se     = unname(new$se),
    ref_se     = as.numeric(ref$se),
    diff_est   = unname(new$estimate) + as.numeric(ref$estimate),
    diff_se    = unname(new$se) - as.numeric(ref$se)
  )
}))
knitr::kable(cmp, digits = 12)
```

## Multi-arm extension

`RobinCar::robincar_mh()` requires a binary treatment column. `RobinCar2`'s
`robin_mh()` extends to all pairwise contrasts in a single call, with a
fully-vectorised matrix-based implementation:

```{r}
robin_mh(
  y_b ~ s1 + s2,
  data = glm_data,
  treatment = treatment ~ pb(s1, s2),
  estimand = "ATE",
  ci_type = "mGR"
)
```

The user can also restrict the comparisons via `pair`:

```{r}
robin_mh(
  y_b ~ s1 + s2,
  data = glm_data,
  treatment = treatment ~ pb(s1, s2),
  pair = against_ref(levels(glm_data$treatment), ref = "pbo")
)
```

## Differences in argument usage

| Concept | `RobinCar::robincar_mh()` | `RobinCar2::robin_mh()` |
|---|---|---|
| Response | `response_col = "y_b"` | LHS of `formula = y_b ~ ...` |
| Treatment | `treat_col = "treatment"` (binary only) | LHS of `treatment = treatment ~ ...` |
| Analysis strata | `strata_cols = c("s1", "s2")` | RHS of `formula` |
| Randomization scheme | implicit | `pb(...)`, `ps(...)`, `sr(...)` on RHS of `treatment` |
| Multi-arm pairwise | not supported | supported via `pair` |
| Output | `MHResult` | `mh_effect` (with `print`/`confint`/`table` methods) |

Both packages compute the same MH estimator, the same Greenland-Robins (GR),
modified Greenland-Robins (mGR), and Sato variance estimators, and the same
ATE additive correction (`nu`). Numerical agreement is verified above and
exercised in the unit tests under `tests/testthat/test-robin_mh.R`.
