---
title: "RobinCar Comparison"
output: rmarkdown::html_document
vignette: >
  %\VignetteIndexEntry{RobinCar Comparison}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
if (!requireNamespace("RobinCar", quietly = TRUE)) {
  install.packages("RobinCar")
}
library(RobinCar2)
library(RobinCar)
```

[RobinCar](https://marlenabannick.com/RobinCar/) was developed as a comprehensive library of existing and future methods on covariate adjustment. RobinCar2 aims to become a standard R package for industry and regulatory agencies to implement covariate adjustment using selected methods with clear documentation and validation for GxP compliance.

*See the American Statistical Association (ASA) BIOP Covariate Adjustment Scientific Working Group (SWG) Software Subteam page [here](https://carswg.github.io/subteam_software.html) for more information.*

As such, RobinCar and RobinCar2 share many methods. Please use this vignette as a guide to the similarities and differences between RobinCar and RobinCar2.

## RobinCar2 and RobinCar produce the same results when run with the same settings on the same data

RobinCar2 has a testing suite that ensures that the outputs are identical between RobinCar and RobinCar2. The comparisons use the `dummy_data` dataset included in RobinCar2. The outputs for the point estimates and variance estimates are compared under a variety of settings, including:

-   Marginal means and treatment contrasts, for the difference, relative risk, odds ratio, and a custom contrast.
-   Under Gaussian, binomial, Poisson, and negative binomial (known and unknown dispersion parameters) working models.
-   Mean model specifications of ANOVA, ANCOVA, and ANHECOVA with up to three covariates.
-   Under simple, Pocock-Simon, and permuted block randomization schemes.

Please see the following testing file for details: <https://github.com/openpharma/RobinCar2/blob/main/tests/testthat/test-validate-robin_glm.R>

## Differences between RobinCar and RobinCar2 argument usage

The interface for users is very similar between RobinCar and RobinCar2, though RobinCar2 has some more streamlined use. Users of RobinCar should find RobinCar2 at least as user-friendly and not have trouble switching between packages.

Here we provide an example of using a *logistic regression working model* with *stratified permuted block randomization*, and the `dummy_data` provided in the package. We output the treatment contrast.

```         
head(dummy_data)
```

### RobinCar

Here is the call in RobinCar, and the output:

```{r, eval=T}
robincar_glm(
  df = dummy_data,
  treat_col = "treatment",
  response_col = "y_b",
  formula = y_b ~ treatment * s1 + covar,
  car_strata_cols = "s1",
  car_scheme = "permuted-block",
  g_family = binomial(link = "logit"),
  contrast_h = "diff"
)
```

### RobinCar2

In RobinCar2, all of the randomization information, including the name of the treatment variable, the stratification variable, and the randomization scheme info is included in the `treatment` argument. This contrasts with RobinCar, where the information is required in `treat_col`, `car_strata_cols`, and `car_scheme`.

```{r, eval=T}
robin_glm(
  data = dummy_data,
  formula = y_b ~ treatment * s1 + covar,
  treatment = treatment ~ pb(s1),
  family = binomial(link = "logit")
)
```

The structure of the output of RobinCar2 more closely mirrors the structure of the output that one would obtain using `glm()`.

## Feature availability in RobinCar v. RobinCar2

Here we list any differences in the features that are available in RobinCar and RobinCar2. This will be continually updated as the packages are updated.

*Last Updated: February 21, 2025*

### Features available in RobinCar that are not available in RobinCar2

#### Variance function

RobinCar only has one option for the variance, which RobinCar2 calls `vcovG`. `vcovG` is the influence function-based variance estimator. RobinCar2 has the option of specifying a `vcov` other than `vcovG`.

Currently, the only other `vcov` that is allowed in RobinCar2 is the sandwich variance, called `vcovHC` and it is only valid for ANCOVA with a linear contrast. Please see the table in the following blogpost from the ASA-BIOP Covariate Adjustment SWG [here](https://carswg.github.io/posts/blog_linear_model.html) for more information about the methodology behind the sandwich variance versus influence-based variance.

### Features available in RobinCar2 that are not available in RobinCar

#### Time-to-event analysis

Currently, RobinCar has functions for time-to-event data analysis and RobinCar2 does not. This is planned for future releases of RobinCar2. Please see [here](https://marlenabannick.com/RobinCar/articles/Robust-Testing-for-Time-to-Event-Data.html) for usage of RobinCar time-to-event methods.
