---
title: "Test on RobinCar2"
author: Gregory Chen
---

## Create dummy data

Before we start our model, first we need to prepare our data.

In the following part we create a simple simulation data.
z1 comes from a bernouli distribution of p = 0.5.
z2 comes from a multinomial distribution of p = (0.4, 0.3, 0.3)
cov1 comes from a normal distribution.
trt is based on stratified block randomization, block size is 6, stratified by z1 and z2.
y is the response, the formula is

exp(y) = 0.2 + 0.5 * (trt = "trt1") + 0.7 * (trt = "trt2") + 1.5 * (z1 = "b") - 1 * (z2 = y) - 0.5 * (z2 = z) + 0.2 * cov

```{r}
n <- 200
block_rand <- function(block = c(0, 0, 1, 1)) {
  function(n) {
    r <- lapply(seq_len(ceiling(n / length(block))), function(i) {
      sample(block)
    })
    unlist(r)[1:n]
  }
}
by_strata <- function(f, strata) {
  ret <- rep(NA, length(strata))
  for (x in split(seq_len(length(strata)), strata)) {
    ret[x] <- f(length(x))
  }
  return(ret)
}
sim_data <- function(n) {
  cov1 <- rnorm(n)
  z1 <- sample(size = n, factor(c("a", "b")), c(0.5, 0.5), replace = TRUE)
  z2 <- sample(size = n, factor(c("x", "y", "z")), c(0.4, 0.3, 0.3), replace = TRUE)
  permute_block <- c(0, 0, 1, 1)
  trt <- by_strata(block_rand(c("trt1", "trt1", "trt2", "trt2", "pbo", "pbo")), interaction(z1, z2))
  trt <- factor(trt, levels = c("pbo", "trt1", "trt2"))
  df <- data.frame(trt, z1, z2, cov1)
  x_mat <- model.matrix(~ trt + z1 + z2 + cov1, data = df)
  coef <- c(0.2, 0.5, 0.7, 1.5, -1, -0.5, 0.2)
  theta <- x_mat %*% coef
  y <- plogis(theta)
  y_bin <- as.integer(runif(n) < y)

  df$y <- y_bin
  df
}

d <- sim_data(500)
```

## Construct the model

In this part, we will create the model based on the data provided.
Let's begin with binary outcome:

```{r}
fit <- glm(y ~ trt:z1 + trt, family = binomial(), data = d)
```

## Obtain the predictions

```{r}
predict_cf <- function(fit, trt, data, ...) {
  UseMethod("predict_cf")
}

predict_cf.glm <- function(fit, trt, data, ...) {
  lvls <- fit$xlevels[[trt]]
  # data has to be a data.frame
  df <- rbind(fit$model, fit$model, fit$model)
  df[[trt]] <- factor(rep(lvls, each = nrow(fit$model)), levels = lvls)
  preds <- predict(fit, type = "response", newdata = df)
  matrix(preds, ncol = length(lvls), dimnames = list(row.names(fit$model), lvls))
}
```

with this `predict_cf` function we already obtain the counterfactual response

```{r}
pred <- predict_cf(fit, "trt")
pred
```

## Obtain the biasness

```{r}
bias <- function(fit, trt, strat, data) {
  UseMethod("bias")
}
bias.glm <- function(fit, trt, strat, data) {
  trt_var <- fit$model[[trt]]
  if (length(strat) != 0) {
    strat_var <- fit$data[, strat]
  } else {
    strat_var <- rep(0L, length(fit$y))
  }
  residuals <- fit$y - fit$fitted.values
  d <- matrix(NA_real_, nrow = length(fit$y), ncol = length(fit$xlevels[[trt]]))
  id_strat <- split(seq_len(length(residuals)), strat_var)
  for (i in id_strat) {
    df <- vapply(split(residuals[i], trt_var[i]), function(xx) mean(xx), FUN.VALUE = 0)
    d[i, ] <- matrix(df, nrow = length(i), ncol = length(df), byrow = TRUE)
  }
  d
}
```

## Ensure the unbiasness

to ensure the unbiasness, the predict_cf need additional argument

```{r}
predict_cf.glm <- function(fit, trt, data, ensure_unbias = TRUE, strata = NULL, ...) {
  lvls <- fit$xlevels[[trt]]
  # data has to be a data.frame
  df <- rbind(fit$model, fit$model, fit$model)
  df[[trt]] <- factor(rep(lvls, each = nrow(fit$model)), levels = lvls)
  preds <- predict(fit, type = "response", newdata = df)
  ret <- matrix(preds, ncol = length(lvls), dimnames = list(row.names(fit$model), lvls))
  if (ensure_unbias) {
    ret <- ret - bias(fit, trt, strata)
  }
  ret
}
```

```{r}
pred <- predict_cf(fit, "trt")
```


## Robust variance

```{r}
adjust_pi <- function(pi_t) {
  diag(pi_t) - pi_t %*% t(pi_t)
}

get_erb <- function(resi, strata, trt, pit, randomization) {
  if (length(strata) == 0) {
    return(0)
  }
  if (randomization %in% c("simple", "pocock-simon")) {
    return(0)
  }
  # Calculate Omega Z under simple
  omegaz_sr <- adjust_pi(pit)
  idx <- split(seq_len(length(resi)), cbind(trt, strata))
  resi_per_strata <- vapply(idx, function(ii) mean(resi[ii]), FUN.VALUE = 0)
  # Calculate strata levels and proportions for
  # the outer expectation

  strata_props <- vapply(idx, length, FUN.VALUE = 0L)
  strata_props <- strata_props / sum(strata_props)
  # Estimate R(B) by first getting the conditional expectation
  # vector for a particular strata (vector contains
  # all treatment groups), then dividing by the pi_t

  rb_z <- resi_per_strata / as.numeric(pit)
  # Compute the R(B)[Omega_{SR} - Omega_{Z_i}]R(B) | Z_i
  # for each Z_i
  strata_levels <- length(resi_per_strata)
  n_trt <- length(pit)
  ind <- matrix(seq_len(strata_levels), byrow = TRUE, ncol = n_trt)
  rb_z_sum <- lapply(
    seq_len(nrow(ind)),
    function(x) rb_z[ind[x, ]] %*% t(rb_z[ind[x, ]]) * sum(strata_props[ind[x, ]])
  )
  rb_z_sum <- Reduce(`+`, rb_z_sum)
  rb_z_sum * omegaz_sr
}

vcov_robin <- function(fit, ...) {
  UseMethod("vcov_robin")
}

are <- function(objs, cls) {
  all(vapply(objs, is, class2 = cls, FUN.VALUE = TRUE))
}

vcov_robin.glm <- function(fit, trt, strat = NULL, preds = predict_cf(fit, trt, strat), sr_decompose = TRUE, randomization = "simple", ...) {
  raw_data <- fit$data
  resi <- residuals(fit, type = "response")
  est <- colMeans(preds)
  var_preds <- var(preds)
  pit <- as.numeric(table(raw_data[[trt]]) / nrow(raw_data))
  trt_lvls <- fit$xlevels[[trt]]

  idx <- split(seq_len(nrow(raw_data)), raw_data[[trt]])
  cov_ymu <- vapply(idx, function(is) stats::cov(fit$y[is], preds[is, ]), FUN.VALUE = rep(0, ncol(preds)))

  vcov_sr <- vapply(idx, function(is) stats::var(fit$y[is]), FUN.VALUE = 0) / pit
  if (sr_decompose) {
    vcov_sr <- vcov_sr + (diag(var_preds) - 2 * diag(cov_ymu)) / pit
  }
  v <- diag(vcov_sr) + cov_ymu + t(cov_ymu) - var_preds
  if (!randomization %in% c("simple", "pocock-simon") && length(strat) > 0) {
    v <- v - get_erb(resi, raw_data[strat], raw_data[[trt]], pit, randomization)
  }
  ret <- v / length(resi)
  dimnames(ret) <- list(trt_lvls, trt_lvls)
  return(ret)
}
vcov_robin(fit, "trt", "z1")
```


## robincar_glm wrapper

```{r}
robincar_glm2 <- function(data, formula, trt, strata, car_scheme, ...) {
  fit <- glm(formula = formula, data = data, ...)
  pred <- predict_cf(fit, trt, ensure_unbias = TRUE)
  var <- vcov_robin(fit, trt = trt, strat = strata, randomization = car_scheme, preds = pred)
  list(
    pred = colMeans(pred),
    var = var
  )
}
robincar_glm(d, formula = y ~ z1:trt + trt, "trt", "z1", "coin")
```

potentially, there can be other arguments just like `RobinCar::robincar_glm`.
The formula can be inferred from the choices of "ANOVA", "ANCOVA", "ANHECOVA", etc. if not provided.

All these functions can be exported for direct usage, and also the wrapper is provided.
